Deployable Probabilistic Programming

### 2019-04-16 paper/paper.tex
better language
code in paper

\documentclass[sigplan,review,10pt,anonymous]{acmart}
\usepackage{url}
\usepackage{hyperref}

@@

	There are two polar views on the role of probabilistic programming.
	One view is that probabilistic programming is a flexible tool for
	specification of statistical models. Proponents of this view
	probabilistic programming as a tool for data science
	practitioners. A typical work flow consists of data acquisition
	and pre-processing, followed by several iterations of exploratory
	model design and testing of inference algorithms. Once a
	sufficiently robust statistical model is obtained, analysis
	results are post-processed, visualized, and occasionally
	integrated into algorithms of a production software system.

@@

	In this work we are concerned with the later view on
	probabilistic programming. Our objective is to pave a path to
	deployment of probabilistic programs in production systems,
	promoting a wider adoption of probabilistic programming as a
	flexible tool for ubiquitous statistical inference.

The language war.

A solution: probabilistic programming without a DSL.

Organization of the paper.

@@

\begin{itemize}
	\item Guidelines for design and implementation of a
		probabilistic programming facility deployable as a part 
		of production software systems.
	\item A reference implementation of deployable probabilistic
		programming facility for the Go programming language.
	\item Design and implementation highlights 
\end{itemize}

@@

A probabilistic model in Infergo is an implementation of an
interface requiring a single method \texttt{Observe} which
accepts a vector (a Go \textit{slice}) of floats, the parameters
to infer, and returns a single float, interpreted as
unnormalized log-likelihood of the posterior distribution.
Implementation of model methods can be written in virtually
unrestricted Go and use any Go libraries.

@@

>>>>

\documentclass[sigplan,review,10pt,anonymous]{acmart}
\usepackage[utf8]{inputenc}
\usepackage{url}
\usepackage{hyperref}
\usepackage{courier}
\usepackage{listings}[language=Golang]

\lstdefinelanguage{Golang}%
  {morekeywords=[1]{package,import,func,type,struct,return,defer,panic,%
     recover,select,var,const,iota,},%
   morekeywords=[2]{string,uint,uint8,uint16,uint32,uint64,int,int8,int16,%
     int32,int64,bool,float32,float64,complex64,complex128,byte,rune,uintptr,%
     error,interface},%
   morekeywords=[3]{map,slice,make,new,nil,len,cap,copy,close,true,false,%
     delete,append,real,imag,complex,chan,},%
   morekeywords=[4]{for,break,continue,range,go,goto,switch,case,fallthrough,if,%
     else,default,},%
   morekeywords=[5]{Println,Printf,Error,Print,},%
   sensitive=true,%
   morecomment=[l]{//},%
   morecomment=[s]{/*}{*/},%
   morestring=[b]',%
   morestring=[b]",%
   morestring=[s]{`}{`},%
}

\lstset{ % add your own preferences
    frame=none,
    basicstyle=\footnotesize\ttfamily,
    numbers=left,
    numbersep=5pt,
    showstringspaces=false,
    tabsize=4,
    language=Golang % this is it !
}

@@

	There are two polar views on the role of probabilistic
	programming.  One view is that probabilistic programming is
	a flexible framework for specification and analysis of
	statistical models.  Proponents of this view perceive
	probabilistic programming as a tool for data science
	practitioners. A typical work flow consists of data
	acquisition and pre-processing, followed by several
	iterations of exploratory model design and testing of
	inference algorithms. Once a sufficiently robust statistical
	model is obtained, analysis results are post-processed,
	visualized, and occasionally integrated into algorithms of a
	production software system.

@@

	In this work we are concerned with the later view on
	probabilistic programming. Our objective is to pave a path
	to deployment of probabilistic programs in production
	systems, promoting a wider adoption of probabilistic
	programming as a flexible tool for ubiquitous statistical
	inference. Most probabilistic programming systems are
	suited, to a certain extent, to be used in either the
	exploratory or the production scenario. However, the
	proliferation of probabilistic programming languages and
	systems~\cite{} one one hand, and scarcity of success
	stories about production use of probabilistic programming on
	the other hand, suggest that there is a need for new ideas
	and approaches to make probabilistic programming models
	available for production use.

	Our attitude differs from the established approach in that
	instead of proposing yet another probabilistic programming
	language, either genuinely different~\cite{} or derived from
	an existing general programming language by extending the
	syntax or changing the semantics~\cite{}, we advocate
	enabling probabilistic inference on models written in a
	general probabilistic programming language, the same one as
	the language used to program the bulk of the system. We
	formulate guidelines for such implementation, and based on
	the guidelines, introduce a probabilistic programming
	facility for the Go programming language. By explaining our
	design choices and implementation techniques, we argue that
	a similar facility can be added to any modern general
	purpose programming language, giving a production system
	modeling and inference capabilities which do not fall short
	from and sometimes exceed those of probabilistic programming
	centric systems with custom languages and runtimes.

@@

	This work brings the following major contributions:

	\begin{itemize}
		\item Guidelines for design and implementation of a
			probabilistic programming facility deployable as a part 
			of production software systems.
		\item A reference implementation of deployable probabilistic
			programming facility for the Go programming language.
		\item Design and implementation highlights outlining
			important choices we made in implementation of the
			probabilistic programming facility, and providing a
			basis for implementation of a similar facility for
			other general programming languages.
	\end{itemize}

@@

A probabilistic model in Infergo is an implementation of an
interface requiring a single method \lstinline{Observe} which
accepts a vector (a Go \textit{slice}) of floats, the parameters
to infer, and returns a single float, interpreted as
unnormalized log-likelihood of the posterior distribution.
Implementation of model methods can be written in virtually
unrestricted Go and use any Go libraries.

@@

A model is defined in it's own package. The model must
implement interface \lstinline{Model} containing a single method
\lstinline{Observe}. \lstinline{Observe} accepts receives a slice of
\lstinline{float64} --- the model parameters --- and returns a
\lstinline{float64} scalar. For probabilistic inference, the
returned value is interpreted as the log-likelihood of the model
parameters. In the model's source code:

\begin{enumerate}
	\item Methods on the type implementing \lstinline{Model} returning a
		single \lstinline{float64} or nothing are differentiated.
	\item Within the methods, the following is differentiated:
		\begin{itemize}
			\item assignments to \lstinline{float64} (including parallel
				assignments if all values are of type
				\lstinline{float64});
			\item returns of \lstinline{float64};
			\item standalone calls to methods on the type implementing
				\lstinline{Model} (apparently called for side  effects on
				the model).
		\end{itemize}
\end{enumerate}

Derivatives do not propagate through a function that is not
an elemental or a call to a model method. If a derivative is
not registered for an elemental, calling the elemental in a
differentiated context will cause a run-time error.

\subsection{Elementals}

Functions are considered elementals (and must have a
registered derivative) if their signature is of kind
\lstinline{func (float64, float64*) float64}; that is, 
one or more non-variadic \lstinline{float64} arguments and
\lstinline{float64} return value. For example, function
\lstinline{func (float64, float64, float64) float64}
is considered elemental, while functions

\begin{itemize}
	\item \lstinline{func (...float64) float64}
	\item \lstinline{func ([]float64) float64}
	\item \lstinline{func (int, float64) float64}
\end{itemize}

are not. Gradients for selected functions from the \lstinline{math}
package are pre-defined (\lstinline{Sqrt}, \lstinline{Exp}, \lstinline{Log}, \lstinline{Pow}, \lstinline{Sin},
\lstinline{Cos}, \lstinline{Tan}). Auxiliary elemental functions with pre-defined
gradients are provided in a separate package.
